---
title: "The Darkside of DBI and dbplyr abstraction"
author: "Matt Sharkey"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

Abstraction makes life easier.  I don't need to know how an enginer works to use a car.  I don't need to understand the chemical properties of metal oxide semiconductors to use a computer.  And I don't need a PhD in   

```{r cars}
summary(cars)
```
## Abstraction - We find virtue between excess and deficiency


```{r, echo=FALSE}
   con <- dbConnect(odbc::odbc(),Driver= myDriver,Server = myServer
                   ,Database = myDatabase,Trusted_Connection='yes')
if(dbExistsTable(con, "yellow_trip_summary_model"))
  {dbRemoveTable(con , 'yellow_trip_summary_model')}
```

```{r}
dbCreateTable(con,'yellow_trip_summary_model',trips_fs)
```

--

```{sql,connection = sqlchunk_connection,echo=FALSE,output.var = "DataTypes" }
Select c.name as Column_Name 
,t.name as Default_Type
,t2.name as Hand_Crafted
,Byte_Diff as Byte_Difference 
from sys.columns c
JOIN sys.types t
ON c.user_type_id = t.user_type_id
JOIN sys.columns cs
ON c.name = cs.name
and cs.object_id = object_id('yellow_trip_summary_partitioned')
JOIN sys.types t2
ON cs.user_type_id = t2.user_type_id
JOIN
(
Select 'VendorID' as name,7 as Byte_Diff UNION ALL
Select 'tpep_pickup_datetime' as name,0 as Byte_Diff UNION ALL
Select 'tpep_dropoff_datetime' as name,0 as Byte_Diff UNION ALL
Select 'passenger_count' as name,7 as Byte_Diff UNION ALL
Select 'trip_distance' as name,6 as Byte_Diff UNION ALL
Select 'RatecodeID' as name,7 as Byte_Diff UNION ALL
Select 'store_and_fwd_flag' as name,0 as Byte_Diff UNION ALL
Select 'PULocationID' as name,6 as Byte_Diff UNION ALL
Select 'DOLocationID' as name,6 as Byte_Diff UNION ALL
Select 'payment_type' as name,7 as Byte_Diff UNION ALL
Select 'fare_amount' as name,3 as Byte_Diff UNION ALL
Select 'extra' as name,3 as Byte_Diff UNION ALL
Select 'mta_tax' as name,3 as Byte_Diff UNION ALL
Select 'tip_amount' as name,3 as Byte_Diff UNION ALL
Select 'tolls_amount' as name,3 as Byte_Diff UNION ALL
Select 'improvement_surcharge' as name,3 as Byte_Diff UNION ALL
Select 'total_amount' as name,3 as Byte_Diff 
) ch
ON c.name = ch.name
Where c.object_id = object_id('yellow_trip_summary_model')
```

```{r , warning= FALSE,echo= FALSE}
knitr::kable(DataTypes,  format="html") %>% 
  kable_styling() %>%
  scroll_box(width = "100%", height = "500px")
```

---

## 67 Bytes Per Row?  So What?

--

- One year of data = 100 million rows

--

- 6.7 GB of wasted space, compounded by:

  * Memory
  * Network
  * Backups
  * Indexes
  * Replication
  
  
---