---
title: "Building Your First Shiny Database Application"
author: "Matthew Sharkey | msharkey3434@gmail.com"
date: "`r Sys.Date()`"
bibliography: ["citation.bib"]
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, echo=FALSE,warning=FALSE,message=FALSE}
options(htmltools.dir.version = FALSE)
library(DBI)
library(dbplyr)
library(tidyverse)
library(kableExtra)
library(microbenchmark)
library(ggplot2)
myDriver <- 'SQL Server'
myServer <- '.\\snapman'
myDatabase <- 'Cab_Demo'
sqlchunk_connection <- dbConnect(odbc::odbc(),
                          Driver= myDriver,
                          Server = myServer,
                          Database = myDatabase,
                          Trusted_Connection='yes')
comp <- Sys.info()
if(comp["nodename"]=="OMC-5CG6151GZF"){
  cab_fileloc <- "C:/Users/MaSharkey/Cab_Data/yellow_tripdata_2018-12.csv"} else {
  cab_fileloc <- "D:/Cab_data/yellow_tripdata_2018-01.csv"
  }
```



# About Me

Matthew Sharkey, MBA, MS-BIA | Data Engineer @ HDR

---

# Agenda

1. RDBMS Use Cases

2. Building and Managing Connections

3. Performance & Optimization

---

# Why use an RDBMS?


- Potential Performance Gains

--

  + Indexing

--

  + Parallelism

--

- Stability

--

- Security

--

- Transactions (ACID)

--

- Change Tracking

--

- But it's not the right tool for every job!

---
## So Help Me Codd!


<br>
<center>
>IT should never forget that technology is a means to an end, and not an end in itself...

--

>...Attempting to force one technology or tool to satisfy a particular need for which another tool is more effective and efficient is like attempting to drive a screw into a wall with a hammer when a screwdriver is at hand: the screw may eventually enter the wall but at what cost?

<br>
<center>
E.F. Codd, et al. 1998. Providing Olap to User-Analysts: An It Mandate. https://bit.ly/2Okyz2H.


---

class: inverse, center, middle

# Building and Managing Connections


---


```{r commonCon, eval=TRUE, warning=FALSE}
myconnection <- DBI::dbConnect(odbc::odbc(),
                          Driver= 'SQL Server',
                          Server = '.\\snapman',
                          Database = 'Cab_Demo',
                          Trusted_Connection='yes')
```

--

```{r execQuery, eval=TRUE, warning=FALSE}

dbGetQuery(myconnection,"Select GetDate()")
dbDisconnect(myconnection)

```

---


## Common Connection Problems

.pull-left[
1. Wrong connection string format 
    
1. Misspelled something

1. Permissions

1. Driver

1. Wrong port number

1. Firewall
]

.pull-right[
<center>
<iframe src="https://giphy.com/embed/Ym0zs6Ms4vgwU" width="380" height="227" frameBorder="0" class="giphy-embed" allowFullScreen></iframe><p><a href="https://giphy.com/gifs/mr-robot-rami-malek-its-so-good-Ym0zs6Ms4vgwU"></a></p>
</center>
]


---

## Manual vs Pool

```{r pool, eval=TRUE,echo=FALSE}
library(pool)
queries <- c("SELECT Getdate()","Select Getdate()","Select Getdate()","Select Getdate()")

dbconnectworkload <- function() {
 for(i in 1:length(queries)){
   con <- dbConnect(odbc::odbc(),Driver= myDriver,Server = myServer
                   ,Database = myDatabase,Trusted_Connection='yes')
  dbGetQuery(con,queries[i])
  dbDisconnect(con)
 }
}  

poolcon <- dbPool(odbc::odbc(),Driver= myDriver,Server = myServer
                  ,Database = myDatabase,Trusted_Connection='yes')

dbpoolworkload <- function() {
 for(i in 1:length(queries)){dbGetQuery(poolcon,queries[i])}
}


```

```{r,eval=FALSE}
  {{dbConnect}}(
    odbc::odbc(),Driver= myDriver,Server = myServer
                ,Database = myDatabase,Trusted_Connection='yes')
  dbGetQuery(con,query)
  dbDisconnect(con)  
```
--
```{r,eval=FALSE}
  {{dbPool}}(
     odbc::odbc(),Driver= myDriver,Server = myServer
                  ,Database = myDatabase,Trusted_Connection='yes')
```
--
```{r poolres, eval=TRUE}
rs <- microbenchmark(dbconnectworkload() 
                    ,dbpoolworkload(),times = 100)
```

---

```{r boxplot,echo=FALSE, fig.align = "center"}
rs<-as.data.frame(rs)
rs$time <- rs$time/1000000
names(rs) <- c("Connection_Type","Milliseconds")
ggplot(rs, aes(x=Connection_Type, y=Milliseconds , fill=Connection_Type)) +
 geom_boxplot(alpha=0.4) +
    theme(text = element_text(size=16))
#
#, fig.height = 4.5, fig.width = 8,
#
```


---

class: inverse, center, middle

# Performance & Optimization

---


## File vs Table Query

<br>
> If your data fits in memory there is no advantage to putting it in a database: it will only be slower and more frustrating.

<br>
Hadley Wickham, RStudio, Edgar Ruiz. 2019. Dbplyr: A dplyr Back End for Databases. https://CRAN.R-project.org/package=dbplyr.

<br>

--

- Frustrating --> Possibly

--

- Slower --> It Depends

---

## File Storage vs Table

```{r,echo=FALSE,message=FALSE}
packages <- list('odbc','DBI','tidyverse','microbenchmark')
aa <-lapply(packages, require, character.only = TRUE,quietly = TRUE)

con <- dbConnect(odbc(),Driver = 'SQL Server',Server = '.\\snapman'
                 , Database = 'Cab_Demo', trusted_connection = TRUE)
#drop index for demo
dm <- dbExecute(con,"DROP INDEX IF EXISTS nc_yellow_trip
              ON [yellow_tripdata_2018-01] ")
```
 Same data but one stored as CSV and one stored as table 
 <div>
 Data source:NYC Yellow Taxi Trips https://on.nyc.gov/2tn71Qq
 
```{r,message=FALSE,echo=FALSE}
if(exists("trips_fs") == FALSE)
{
trips_fs <- read_csv(cab_fileloc)
trips_db <- tbl(con, "yellow_tripdata_2018-01")
}
```

```{r,message=FALSE,eval=FALSE}
trips_fs <- read_csv(cab_fileloc)
trips_db <- tbl(con, "yellow_tripdata_2018-01")
```
--
```{r,message=FALSE}
fs <- function(){csvquery<- trips_fs %>% 
    filter(tpep_dropoff_datetime >= '2018-01-02 07:28:00'
          ,tpep_dropoff_datetime <= '2018-01-02 07:30:00') %>%
    summarise(pcount= n())
  rs<- suppressMessages(compute(csvquery))}
```
--
```{r}
db <- function(){sqltable <- trips_db %>% 
    filter(tpep_dropoff_datetime >= '01-02-2018 13:28'
           ,tpep_dropoff_datetime <= '01/02/2018 13:30')  %>%
    summarise(pcount= n())
  rs1<-suppressMessages(compute(sqltable))}
```


---

## Benchmarking

File storage is faster


```{r boxplot2,echo=FALSE, fig.align = "center"}
rs<-microbenchmark(db(),fs(),times = 100)
rs<-as.data.frame(rs)
rs$time <- rs$time/1000000
ggplot(rs, aes(x=expr, y=time, fill=expr)) +
 geom_boxplot(alpha=0.4) +
    theme(text = element_text(size=20))

```



---

## But...

Indexing makes a difference

```{r,echo=FALSE,message=FALSE}
ms<-dbExecute(con,
'CREATE NONCLUSTERED INDEX nc_yellow_trip
 ON [yellow_tripdata_2018-01](tpep_dropoff_datetime)'
 )
```

```{r,eval=FALSE}
dbExecute(con,
'CREATE NONCLUSTERED INDEX nc_yellow_trip
 ON [yellow_tripdata_2018-01](tpep_dropoff_datetime)'
 )
```

--

```{r boxplot3,echo=FALSE, fig.align = "center"}
rs<-microbenchmark(db(),fs(),times = 100)

rs<-as.data.frame(rs)
rs$time <- rs$time/1000000
ggplot(rs, aes(x=expr, y=time, fill=expr)) +
 geom_boxplot(alpha=0.4) +
    theme(text = element_text(size=20))

```


```{r,echo=FALSE}
dbDisconnect(con)
```


--
<br>
The tables have turned.

---


## Indexes - The Keys to Query Performance


Good columns for indexing

--

- Remember FOC

--

F - Filtering
```{r,eval=FALSE}
mtcars %>% filter(cyl == 4)
```
--

O- Ordering
```{r,eval=FALSE}
mtcars %>% filter(cyl == 4) %>% arrange(disp)
```
--

C- Covering (Any remaining column in the SELECT)
```{r,eval=FALSE}
mtcars %>% filter(cyl == 4) %>% arrange(disp) %>% select(mpg)
```
--
```{sql,connection = sqlchunk_connection ,eval=FALSE}
CREATE INDEX ix_mtcars ON mtcars(cyl,disp)
INCLUDE (mpg)
```
---

## Best index for this query?

```{r,eval= FALSE}
trips_db %>%
      mutate(Trip_Date = as.Date(tpep_dropoff_datetime))  %>% 
      filter(tpep_dropoff_datetime >= '01/01/2018',
             tpep_dropoff_datetime < '01/04/2018',
             trip_distance >= 0,
             trip_distance <= 5)  %>%
      group_by(Trip_Date) %>%
      summarise(Trip_Count= n())

```


--

- Per FPOC, start with tpep_dropoff_datetime and trip_distance

--

```{sql,eval=FALSE,connection=sqlchunk_connection}
CREATE INDEX ix_yellow_cab 
ON yellow_trip(tpep_dropoff_datetime,trip_distance)
```

--

- Test performance before and after index

- Indexing trade-off = faster reads, slower writes

---

## Demo

- Yellow Cab Dashboard


---


# Recap

- Use the right tool for the job

--

- Troubleshooting connections

--

- Performance

--

Presentation write up on blog [www.hintank.com](https://www.hinttank.com/)

--

Demo Application on github https://github.com/msharkey/shinybuilder


---

class: center, middle

# Thank You


